---
title: "Statistical Computing Homework 4"
# subtitle: "possible subtitle goes here"
author:
  - Jieying Jiao
date: "`r format(Sys.time(), '%d %B %Y')`"
documentclass: article
papersize: letter
fontsize: 11pt
bibliography: template.bib
biblio-style: asa
keywords: Template, R Markdown, bookdown, Data Lab
# keywords set in YAML header here only go to the properties of the PDF output
# the keywords that appear in PDF output are set in latex/before_body.tex
output:
  bookdown::pdf_document2
  bookdown::html_document2
abstract: |
    This is Jieying Jiao's homework 4 for statistical computing, fall 2018.
---

```{r setup, echo = FALSE, message = FALSE, warning = FALSE}
## some utility functions, see the source code for details
source("utils_template.R")

## specify the packages needed
pkgs <- c("DT", "leaflet", "splines2", "webshot")
need.packages(pkgs)

## external data can be read in by regular functions,
## such as read.table or load

## for latex and html output
isHtml <- knitr::is_html_output()
isLatex <- knitr::is_latex_output()
latex <- ifelse(isLatex, '\\LaTeX\\', 'LaTeX')

## specify global chunk options
knitr::opts_chunk$set(fig.width = 5, fig.height = 4, dpi = 300,
                      out.width = "90%", fig.align = "center")

```


# Exercise 3.3.2

```{r HW4_1, warning=FALSE}
##' define the loglikelihood function
loglik.my0 <- function(theta, sample) {
  n <- length(sample)
  if (sum(sample >=0 & sample <= 2*pi) < n) {
    print("sample is out of range")
  } else if(theta < -pi | theta > pi) {
    print("theta is out of range")
  } else {
   l <- sum(log(1-cos(sample-theta))) - n * log(2*pi)
   return(l) 
  }
}
loglik.my <- function(theta, sample) {
  l <- sapply(theta, FUN = loglik.my0, sample = sample)
  l
}

s <- c(3.91, 4.85, 2.28, 4.06, 3.70, 4.04, 5.46, 3.53, 2.28, 1.96,
       2.53, 3.88, 2.22, 3.47, 4.82, 2.46, 2.99, 2.54, 0.52)
curve(loglik.my(x, sample = s), -pi, pi)
```

\begin{align*}
E(X|\theta) &= \int_0^{2\pi}x\frac{1-\cos(x-\theta)}{2\pi}dx\\
            &= \frac{1}{2\pi} \int_0^{2\pi}xdx - \frac{1}{2\pi}\int_0^{2\pi}
            x\cos(x-\theta)dx\\
            &= \left.\frac{1}{2\pi} \times \frac{1}{2}\right\vert_0^{2\pi} - 
            \frac{1}{2\pi}\int_0^{2\pi}xd\sin(x-\theta)\\
            &= \pi - \frac{1}{2\pi} \left[\left.x\sin(x-\theta)\right\vert_0^
            {2\pi}-\int_0^{2\pi}\sin(x-\theta)dx\right]\\
            &= \pi - \frac{1}{2\pi} \left[2\pi\sin(2\pi-\theta)+\cos(x-\theta)
            \left.\right\vert_0^{2\pi}\right]\\
            &= \pi - \frac{1}{2\pi} \left[-2\pi\sin(\theta)\right]\\
            &= \pi + \sin(\theta)\\
            &= \bar{X}_n\\
\end{align*}
$$\Rightarrow\tilde{\theta}_n = \arcsin(\bar{X}_n - \pi)$$
```{r HW4_2, warning=FALSE}
library(pracma)
theta_0 <- asin(mean(s) - pi)
##' define derivitive of log-likelihood function
dev.loglik0 <- function(theta, sample) {
  dev.l <- sum(sin(theta-sample)/(1-cos(theta-sample)))
  dev.l
}
dev.loglik <- function(theta, sample) {
  dev.l <- sapply(theta, FUN = dev.loglik0, sample = sample)
}

x1 <- newtonRaphson(fun = dev.loglik, x0 = theta_0, sample = s)$root
x2 <- newtonRaphson(fun = dev.loglik, x0 = -2.7, sample = s)$root
x3 <- newtonRaphson(fun = dev.loglik, x0 = 2.7, sample = s)$root
x1
x2
x3
```

Newton-Raphson method gives 0.0031 as MLE when MOM 0.095 is initial value, 
-2.669 as MLE when -2.7 is initial value, 2.848 as MLE when 2.7 is initial value.

```{r HW4_3, warning=FALSE}
n <- 200
init <- seq(-pi, pi, length.out = n)
root.newton <- rep(0, n)
for (i in 1:n) root.newton[i] <- newtonRaphson(fun = dev.loglik, x0 = init[i], 
                                               sample = s)$root
plot(init, root.newton, type = "l", xlab = "initial value")
```

# Exercise 3.3.3

```{r HW4_3, warning=FALSE}
beetles <- data.frame(
    days    = c(0,  8,  28,  41,  63,  69,   97, 117,  135,  154),
    beetles = c(2, 47, 192, 256, 768, 896, 1120, 896, 1184, 1024))

```